Stability concerns in random matrices

### 2018-03-28 random_matrix_stability.Rmd
Code examples up

@@

A lot is going to change.

@@

@@

**Table 1:** Stability results from 100 million simulations of random interaction matrices for 2-15 species ($S$). Columns 2 and 3 report the number of stable and unstable communities in the absence of variation in species generation time (all $\gamma = 1$). Columns 4 and 5 report the number of unstable communities stabilised or stable communities destabilised by $Var(\gamma)$, respectively. Column 6 reports the total number of communities that are stable given $Var(\gamma)$ (compare to column 3).

@@

@@

**Figure 1:** Stability of random communities when generation time ($\gamma$) varies and when it does not for 2-15 species (x-axis). Bars compare the $\ln$ total number of communities found to be stable out of 100 million before (dark grey) and after (light grey) $Var(\gamma)$ is added to a community (left y-axis). Points show the proportion of these random communities in which stability is a direct consequence of $Var(\gamma)$.

>>>>

**NEW AUTHORS NOTE:** After analysing the new results, I am thinking that a different angle is needed for a strong publication. To recap what has happened:

- I started analysing random communities of simulated species using matrices and a mathematical trick similar to that used in @Patel2018 in attempt to see if variation in generation time among species will stabilise or destabilise large communities.
- I initially found variation in generation time to be highly stabilising, and, in particular, as species richness increased, the overwhelming number of stable random communties were stable *due to variation in generation time* (and would otherwise be unstable). Results are summarised in [this figure](#old_fig_1) and [this table](#old_tab_1).
- I then read the very recent paper by @Dougoud2018, who showed that the classic stability criteria used by @May1972 and @Allesina2012 (among others) have been flawed. Some "stable" communities in these studies are stable (i.e., return to equilibrium values if perturbed) at *negative* species densities. @Dougoud2018 demonstrate how to check whether or not random communities are "feasible", meaning that they are stable and have species densities that are all positive (and hence, sensible ecological communities).
- I applied the criteria of @Dougoud2018 to my random communities under the same parameter values that they and others have used; I also looked at communities characterised only by competitive, mutualistic, and predator-prey interactions.
- Strangely, variation in generation time has the exact *opposite* effect in random, competitive, and predator-prey communities; variation in generation time therefore *increases* stability but *decreases* feasibility (i.e., variation in generation time shifts communities toward being "stable" with negative species densities and away from being "stable" with positive species densities). The overall effect is therefore to make real ecological communities less stable.
- I am still working out what this means conceptually by multiplying each row of the matrix by a single number, the shape of the matrix is somehow being changed to decrease the eigenvalues.
- The exception to this is mutualistic communities -- variation in generation time had zero effect on mutualist community stability.
- **I am tempted to broaden these results to make a more general point about complex dynamic systems**: "Variation in process time increases stability but decreases feasibility in complex systems", with a focus on biological systems. There might be some types of systems where negative values are actually meaningful (e.g., phenotypes in evolving systems, gene regulatory networks? Suggestions very welcome), just not in ecological ones where species densities obviously cannot be negative.

The plot below shows the story for the effect of variation in generation time on stability and feasibility.

And here's the same plot with the axes shifted a bit.

```{r, echo = FALSE, fig.height = 10, fig.width = 10}
random      <- read.csv(file = "sim_results/random_community.csv");
competition <- read.csv(file = "sim_results/competitor_community.csv");
mutualism   <- read.csv(file = "sim_results/mutualist_community.csv");
pred_prey   <- read.csv(file = "sim_results/predator_prey_community.csv");
# ==============================================================================
par(mfrow = c(2, 2));
# ==============================================================================
dat      <- random;
Sp       <- dat[,1];
stab_nov <- dat[,3] + dat[,4];
stab_var <- dat[,3] + dat[,5];
feas_nov <- dat[,7] + dat[,8];
feas_var <- dat[,7] + dat[,9];
plot(x = Sp, y = stab_nov / 100000, type = "l", lwd = 2, ylim = c(0,1),
     cex.lab = 1.5, cex.axis = 1.5, xlab = "Species Number (S)",
     ylab = "Proportion stable or feasible", xlim = c(0, 30));
points(x = Sp, y = stab_var / 100000, type = "l", lwd = 2, lty = "dashed");
points(x = Sp, y = feas_nov / 100000, type = "l", lwd = 2, col = "red");
points(x = Sp, y = feas_var / 100000, type = "l", lwd = 2, col = "red",
       lty = "dashed");
legend(x = 12, y = 0.9, col = c("black", "black", "red", "red"), cex = 1.2,
       legend = c(expression(paste("Stable, no Var(",gamma,")")),
                             expression(paste("Stable, Var(",gamma,")")),
                             expression(paste("Feasible, no Var(",gamma,")")),
                             expression(paste("Feasible, Var(",gamma,")"))),
       lty = c("solid", "dashed", "solid", "dashed"), lwd = 2);
text(x = 17, y = 0.98, labels = "Random", cex = 2);
# ==============================================================================
dat      <- competition;
stab_nov <- dat[,3] + dat[,4];
stab_var <- dat[,3] + dat[,5];
feas_nov <- dat[,7] + dat[,8];
feas_var <- dat[,7] + dat[,9];
plot(x = Sp, y = stab_nov / 100000, type = "l", lwd = 2, ylim = c(0,1),
     cex.lab = 1.5, cex.axis = 1.5, xlab = "Species Number (S)",
     ylab = "Proportion stable or feasible", xlim = c(0, 30));
points(x = Sp, y = stab_var / 100000, type = "l", lwd = 2, lty = "dashed");
points(x = Sp, y = feas_nov / 100000, type = "l", lwd = 2, col = "red");
points(x = Sp, y = feas_var / 100000, type = "l", lwd = 2, col = "red",
       lty = "dashed");
text(x = 15, y = 0.98, labels = "Competitive", cex = 2);
# ==============================================================================
dat      <- mutualism;
stab_nov <- dat[,3] + dat[,4];
stab_var <- dat[,3] + dat[,5];
feas_nov <- dat[,7] + dat[,8];
feas_var <- dat[,7] + dat[,9];
plot(x = Sp, y = stab_nov / 100000, type = "l", lwd = 2, ylim = c(0,1),
     cex.lab = 1.5, cex.axis = 1.5, xlab = "Species Number (S)",
     ylab = "Proportion stable or feasible", xlim = c(0, 10));
points(x = Sp, y = stab_var / 100000, type = "l", lwd = 2, lty = "dashed");
points(x = Sp, y = feas_nov / 100000, type = "l", lwd = 2, col = "red");
points(x = Sp, y = feas_var / 100000, type = "l", lwd = 2, col = "red",
       lty = "dashed");
text(x = 6, y = 0.98, labels = "Mutualist", cex = 2);
# ==============================================================================
dat      <- pred_prey;
stab_nov <- dat[,3] + dat[,4];
stab_var <- dat[,3] + dat[,5];
feas_nov <- dat[,7] + dat[,8];
feas_var <- dat[,7] + dat[,9];
plot(x = Sp, y = stab_nov / 100000, type = "l", lwd = 2, ylim = c(0,1),
     cex.lab = 1.5, cex.axis = 1.5, xlab = "Species Number (S)",
     ylab = "Proportion stable or feasible");
points(x = Sp, y = stab_var / 100000, type = "l", lwd = 2, lty = "dashed");
points(x = Sp, y = feas_nov / 100000, type = "l", lwd = 2, col = "red");
points(x = Sp, y = feas_var / 100000, type = "l", lwd = 2, col = "red",
       lty = "dashed");
text(x = 60, y = 0.98, labels = "Predator-prey", cex = 2);
# ==============================================================================
```

We can delve a bit deeper into this mess by looking at the distribution of eigenvalues in sample communities that are unstable, stable, destabilised, or stabilised. The plot below shows these examples for communities where species number $S = 20$. This gives the same type of plots as shown in @Allesina2012. Black points show eigenvalues before variation in generation time is added, and red points show eigenvalues after variation in generation time is added. The shift is not typically drastic, but sometimes causes the sign of the real components (x-axis) to flip, thereby potentially affecting stability. Immediately below shows plots for a random community; where not points are plot in any of the below, no examples exist.

```{r, echo = FALSE}
rand_gen_var <- function(max_sp, iters, int_type = 0, rmx = 10, eps_max = 999){
    tot_res <- NULL;
    fea_res <- NULL;
    for(i in 2:max_sp){
        nn             <- i;
        A1_stt         <- 0;
        A2_stt         <- 0;
        A1_fet         <- 0;
        A2_fet         <- 0;
        iter           <- iters;
        tot_res[[i-1]] <- matrix(data = 0, nrow = iter, ncol = 2);
        fea_res[[i-1]] <- matrix(data = 0, nrow = iter, ncol = 2);
        while(iter > 0){
            r_vec    <- runif(n = nn, min = 1, max = rmx);
            A1_dat   <- rnorm(n = nn * nn, mean = 0, sd = 0.4);
            A1       <- matrix(data = A1_dat, nrow = nn, ncol = nn);
            A1       <- species_interactions(mat = A1, type = int_type);
            diag(A1) <- -1;
            epsil    <- runif(n = nn, min = 1, max = eps_max);
            eps_dat  <- rep(x = epsil, times = nn);
            eps_mat  <- matrix(data = epsil, nrow = nn, ncol = nn, 
                               byrow = FALSE);
            avg_mat  <- matrix(data = mean(1:eps_max), nrow = nn, ncol = nn, 
                               byrow = FALSE);
            A2       <- A1 * eps_mat;
            A1       <- A1 * avg_mat;
            A1_stb   <- max(Re(eigen(A1)$values)) < 0;
            A2_stb   <- max(Re(eigen(A2)$values)) < 0;
            A1_fea   <- min(-1*solve(A1) %*% r_vec) > 0;
            A2_fea   <- min(-1*solve(A2) %*% r_vec) > 0;
            if(A1_stb == TRUE){
                tot_res[[i-1]][iter, 1] <- 1;
            }
            if(A2_stb == TRUE){
                tot_res[[i-1]][iter, 2] <- 1;
            }
            if(A1_fea == TRUE){
                fea_res[[i-1]][iter, 1] <- 1;
            }
            if(A2_fea == TRUE){
                fea_res[[i-1]][iter, 2] <- 1;
            }
            iter    <- iter - 1;
        }
        print(i);
    }
    all_res <- summarise_randmat(tot_res = tot_res, fea_res = fea_res);
    return(all_res);
}

species_interactions <- function(mat, type = 0){
    if(type == 1){
        mat[mat > 0] <- -1*mat[mat > 0];
    }
    if(type == 2){
        mat[mat < 0] <- -1*mat[mat < 0];
    }
    if(type == 3){
        for(i in 1:dim(mat)[1]){
            for(j in 1:dim(mat)[2]){
                if(mat[i, j] * mat[j, i] > 0){
                    mat[j, i] <- -1 * mat[j, i];
                }
            }
        }
    }
    return(mat);
}

summarise_randmat <- function(tot_res, fea_res){
    sims    <- length(tot_res);
    all_res <- matrix(data = 0, nrow = sims, ncol = 9);
    for(i in 1:sims){
        unstables <- tot_res[[i]][,1] == FALSE & tot_res[[i]][,2] == FALSE;
        stables   <- tot_res[[i]][,1] == TRUE  & tot_res[[i]][,2] == TRUE;
        unstabled <- tot_res[[i]][,1] == TRUE  & tot_res[[i]][,2] == FALSE;
        stabled   <- tot_res[[i]][,1] == FALSE & tot_res[[i]][,2] == TRUE;
        non_feas  <- fea_res[[i]][,1] == FALSE & fea_res[[i]][,2] == FALSE;
        feasibl   <- fea_res[[i]][,1] == TRUE  & fea_res[[i]][,2] == TRUE;
        unfeased  <- fea_res[[i]][,1] == TRUE  & fea_res[[i]][,2] == FALSE;
        feased    <- fea_res[[i]][,1] == FALSE & fea_res[[i]][,2] == TRUE;
        all_res[i, 1] <- i + 1;
        all_res[i, 2] <- sum(unstables);
        all_res[i, 3] <- sum(stables);
        all_res[i, 4] <- sum(unstabled);
        all_res[i, 5] <- sum(stabled);
        all_res[i, 6] <- sum(non_feas);
        all_res[i, 7] <- sum(feasibl);
        all_res[i, 8] <- sum(unfeased);
        all_res[i, 9] <- sum(feased);
    }
    return(all_res);
}

get_rand_eigs <- function(int_type = 0, iter = 1000){
    nn             <- 20;
    eigres         <- matrix(data = 0, ncol = 82, nrow = iter);
    eigres[,1]     <- int_type;
    while(iter > 0){
        r_vec     <- runif(n = nn, min = 1, max = 1000);
        A1_dat    <- rnorm(n = nn * nn, mean = 0, sd = 0.4);
        A1        <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        A1        <- species_interactions(mat = A1, type = int_type);
        diag(A1)  <- -1;
        epsil     <- runif(n = nn, min = 1, max = 999);
        eps_dat   <- rep(x = epsil, times = nn);
        eps_mat   <- matrix(data = epsil, nrow = nn, ncol = nn, 
                            byrow = FALSE);
        avg_mat   <- matrix(data = mean(1:999), nrow = nn, ncol = nn, 
                            byrow = FALSE);
        A2        <- A1 * eps_mat;
        A1        <- A1 * avg_mat;
        mat1_e    <- eigen(A1)$values;
        mat2_e    <- eigen(A2)$values;
        real_mat1 <- Re(mat1_e);
        imag_mat1 <- Im(mat1_e);
        real_mat2 <- Re(mat2_e);
        imag_mat2 <- Im(mat2_e);
        stab1     <- max(real_mat1) < 0;
        stab2     <- max(real_mat2) < 0;
        if(stab1 == FALSE & stab2 == FALSE){
            eigres[iter,2] <- 1;
        }
        if(stab1 == TRUE & stab2 == TRUE){
            eigres[iter,2] <- 2;
        }
        if(stab1 == TRUE & stab2 == FALSE){
            eigres[iter,2] <- 3;
        }
        if(stab1 == FALSE & stab2 == TRUE){
            eigres[iter,2] <- 4;
        }
        eigres[iter,3:22]  <- real_mat1;
        eigres[iter,23:42] <- imag_mat1;
        eigres[iter,43:62] <- real_mat2;
        eigres[iter,63:82] <- imag_mat2;
        iter    <- iter - 1;
    }
    return(eigres);
}

plot_eig_set <- function(eigres){
    par(mfrow = c(2,2), mar = c(5, 5, 1, 1));
    x0 <- min(cbind(eigres[,3:22], eigres[,43:62]));
    x1 <- max(cbind(eigres[,3:22], eigres[,43:62]));
    plot(x = 0, y = 0, type = "n", xlim = c(x0, x1), ylim = c(-1500, 1500),
         xlab = "Real component", ylab = "Imaginary component", cex.lab = 1.25);
    abline(h = 0, lty = "dotted", lwd = 0.8);
    abline(v = 0, lty = "dotted", lwd = 0.8);
    if(1 %in% eigres[,2]){
        row <- which(eigres[,2] == 1)[1];
        m1r <- eigres[row,3:22];
        m1i <- eigres[row,23:42];
        m2r <- eigres[row,43:62];
        m2i <- eigres[row,63:82];
        points(x = m1r, y = m1i, pch = 20, cex = 1.5, col = "black");
        points(x = m2r, y = m2i, pch = 20, cex = 1.5, col = "red");
    }
    text(x = x0 + (x1 - x0)*0.2, y = 1500, "Unstable", cex = 1.5);
    plot(x = 0, y = 0, type = "n", xlim = c(x0, x1), ylim = c(-1500, 1500),
         xlab = "Real component", ylab = "Imaginary component", cex.lab = 1.25);
    abline(h = 0, lty = "dotted", lwd = 0.8);
    abline(v = 0, lty = "dotted", lwd = 0.8);
    if(2 %in% eigres[,2]){
        row <- which(eigres[,2] == 2)[1];
        m1r <- eigres[row,3:22];
        m1i <- eigres[row,23:42];
        m2r <- eigres[row,43:62];
        m2i <- eigres[row,63:82];
        points(x = m1r, y = m1i, pch = 20, cex = 1.5, col = "black");
        points(x = m2r, y = m2i, pch = 20, cex = 1.5, col = "red");
    }
    text(x = x0 + (x1 - x0)*0.2, y = 1500, "Stable", cex = 1.5);
    plot(x = 0, y = 0, type = "n", xlim = c(x0, x1), ylim = c(-1500, 1500),
         xlab = "Real component", ylab = "Imaginary component", cex.lab = 1.25);
    abline(h = 0, lty = "dotted", lwd = 0.8);
    abline(v = 0, lty = "dotted", lwd = 0.8);
    if(3 %in% eigres[,2]){
        row <- which(eigres[,2] == 3)[1];
        m1r <- eigres[row,3:22];
        m1i <- eigres[row,23:42];
        m2r <- eigres[row,43:62];
        m2i <- eigres[row,63:82];
        points(x = m1r, y = m1i, pch = 20, cex = 1.5, col = "black");
        points(x = m2r, y = m2i, pch = 20, cex = 1.5, col = "red");
    }
    text(x = x0 + (x1 - x0)*0.2, y = 1500, "Destabilised", cex = 1.5);
    plot(x = 0, y = 0, type = "n", xlim =c(x0, x1), ylim = c(-1500, 1500),
         xlab = "Real component", ylab = "Imaginary component", cex.lab = 1.25);
    abline(h = 0, lty = "dotted", lwd = 0.8);
    abline(v = 0, lty = "dotted", lwd = 0.8);
    if(4 %in% eigres[,2]){
        row <- which(eigres[,2] == 4)[1];
        m1r <- eigres[row,3:22];
        m1i <- eigres[row,23:42];
        m2r <- eigres[row,43:62];
        m2i <- eigres[row,63:82];
        points(x = m1r, y = m1i, pch = 20, cex = 1.5, col = "black");
        points(x = m2r, y = m2i, pch = 20, cex = 1.5, col = "red");
    }
    text(x = x0 + (x1 - x0)*0.2, y = 1500, "Stabilised", cex = 1.5);
}
```

```{r, echo = FALSE, fig.height = 10, fig.width = 10}
eres <- get_rand_eigs(int_type = 0, iter = 10000);
plot_eig_set(eres);
```

Below shows plots for competitive communities.

```{r, echo = FALSE, fig.height = 10, fig.width = 10}
eres <- get_rand_eigs(int_type = 1, iter = 10000);
plot_eig_set(eres);
```

Below shows plots for mutualist communities.

```{r, echo = FALSE, fig.height = 10, fig.width = 10}
eres <- get_rand_eigs(int_type = 2, iter = 10000);
plot_eig_set(eres);
```

Below shows plots for predator-prey communities.

```{r, echo = FALSE, fig.height = 10, fig.width = 10}
eres <- get_rand_eigs(int_type = 3, iter = 10000);
plot_eig_set(eres);
```

What I would like to explore is some sort of pattern in these points that links to the previous plot showing probability of stabilisation versus destabilisation of generation time, but I'm stuck for now. What I can do is look at stable versus feasible systems; the latter seem pretty much the same, except that real eigenvalues in feasible systems are all lower values. Below shows one example for a predator-prey community of $S = 20$.

```{r, echo = FALSE, fig.height = 6, fig.width = 10}
get_rand_feas <- function(int_type = 0, iter = 1000){
    nn             <- 20;
    eigres         <- matrix(data = 0, ncol = 42, nrow = iter);
    eigres[,1]     <- int_type;
    while(iter > 0){
        r_vec     <- runif(n = nn, min = 1, max = 10);
        A1_dat    <- rnorm(n = nn * nn, mean = 0, sd = 0.4);
        A1        <- matrix(data = A1_dat, nrow = nn, ncol = nn);
        A1        <- species_interactions(mat = A1, type = int_type);
        diag(A1)  <- -1;
        avg_mat   <- matrix(data = mean(1:999), nrow = nn, ncol = nn, 
                            byrow = FALSE);
        A1        <- A1 * avg_mat;
        mat1_e    <- eigen(A1)$values;
        real_mat1 <- Re(mat1_e);
        imag_mat1 <- Im(mat1_e);
        stab1     <- max(real_mat1) < 0;
        A1_fea    <- min(-1*solve(A1) %*% r_vec) > 0;
        if(stab1 == TRUE & A1_fea == FALSE){
            eigres[iter,2] <- 1;
        }
        if(stab1 == TRUE & A1_fea == TRUE){
            eigres[iter,2] <- 2;
        }
        eigres[iter,3:22]  <- real_mat1;
        eigres[iter,23:42] <- imag_mat1;
        iter    <- iter - 1;
    }
    return(eigres);
}

plot_feas_set <- function(eigres){
    par(mfrow = c(1,2), mar = c(5, 5, 1, 1));
    x0 <- min(eigres[,3:22]);
    x1 <- max(eigres[,3:22]);
    plot(x = 0, y = 0, type = "n", xlim = c(x0, x1), ylim = c(-1500, 1500),
         xlab = "Real component", ylab = "Imaginary component", cex.lab = 1.25);
    abline(h = 0, lty = "dotted", lwd = 0.8);
    abline(v = 0, lty = "dotted", lwd = 0.8);
    if(1 %in% eigres[,2]){
        row <- which(eigres[,2] == 1)[1];
        m1r <- eigres[row,3:22];
        m1i <- eigres[row,23:42];
        points(x = m1r, y = m1i, pch = 20, cex = 1.5, col = "black");
    }
    text(x = x0 + (x1 - x0)*0.2, y = 1500, "Stable", cex = 1.5);
    plot(x = 0, y = 0, type = "n", xlim = c(x0, x1), ylim = c(-1500, 1500),
         xlab = "Real component", ylab = "Imaginary component", cex.lab = 1.25);
    abline(h = 0, lty = "dotted", lwd = 0.8);
    abline(v = 0, lty = "dotted", lwd = 0.8);
    if(2 %in% eigres[,2]){
        row <- which(eigres[,2] == 2)[1];
        m1r <- eigres[row,3:22];
        m1i <- eigres[row,23:42];
        points(x = m1r, y = m1i, pch = 20, cex = 1.5, col = "black");
    }
    text(x = x0 + (x1 - x0)*0.2, y = 1500, "Feasible", cex = 1.5);
}
fea_test <- get_rand_feas(iter = 10000, int_type = 3);
plot_feas_set(fea_test)
```

@@

@@

<a name = "old_tab_1">**Table 1:**</a> Stability results from 100 million simulations of random interaction matrices for 2-15 species ($S$). Columns 2 and 3 report the number of stable and unstable communities in the absence of variation in species generation time (all $\gamma = 1$). Columns 4 and 5 report the number of unstable communities stabilised or stable communities destabilised by $Var(\gamma)$, respectively. Column 6 reports the total number of communities that are stable given $Var(\gamma)$ (compare to column 3).

@@

@@

<a name = "old_fig_1">**Figure 1:**</a> Stability of random communities when generation time ($\gamma$) varies and when it does not for 2-15 species (x-axis). Bars compare the $\ln$ total number of communities found to be stable out of 100 million before (dark grey) and after (light grey) $Var(\gamma)$ is added to a community (left y-axis). Points show the proportion of these random communities in which stability is a direct consequence of $Var(\gamma)$.

@@
